The Evolution of Phenotypically Invariant Genetic Networks

### 2017-11-17 kalman_walk_paper.tex
overleaf edits

\begin{abstract}
It is known that even if a species' phenotype has remained unchanged over evolutionary time,
the underlying mechanism may have changed,
since distinct molecular pathways can realize identical phenotypes.
In this paper, we use quantitative genetics and linear systems theory to study
how a gene network underlying a conserved phenotype
evolves as genetic drift of small mutational tweaks to the molecular pathways
cause a population to explore the set of mechanisms with identical phenotypes.
In this setting we treat organisms as ``black boxes'' for which the environment provides input
and the phenotype is the output,
and there exists an exact characterization of the set of all mechanisms that give the same input--output relationship. 
We show that in this situation, there is never a unique architecture for any phenotype
and that the evolutionary exploration of these distinct and mutationally connected mechanisms
can lead to the rapid accumulation of hybrid incompatibilities between allopatric populations.
We estimate that in reasonably numerous species,
this process could thus the formation of new species
over a number of generations proportional to the effective population size.
\end{abstract}

@@

\paragraph{A new mechanism underlying Haldane's rule}
The faster rate of fitness decay observed in $F_2$ relative to $F_1$ hybrids is consistent with Haldane's rule: the observation that if only one hybrid sex is sterile or inviable it is likely the heterogametic sex (e.g. the male XY) \citep{haldane1922sex}.
If the regulatory coefficients for one system are shared between the sex and one or more autosomal chromosomes, and dosage compensation mechanisms are present (e.g. the homogametic sex silences or inactivates the expression of one of her two copies), $F_1$ males are effectively equivalent to purely autosomal-system $F_2$s. This is because incompatibilities between loci on the sex chromome with autosomal loci in $F_1$ males are less likely to be balanced.
\footnote{For example, in a 2 gene network where the first gene resides on an autosome and the second gene on the X chromosome, the male with system $\left( \left[ \begin{smallmatrix} a_1 & a_2 \end{smallmatrix} \right], \left[ \begin{smallmatrix} a_1 & a_2 \\ X_1 & X_2 \end{smallmatrix} \right] \right)$ hybridizing with the homozygous female $\left( \left[\begin{smallmatrix} \bar a_1 & \bar a_2 \\ \bar X_1 & \bar X_2 \end{smallmatrix}\right] \right)$ will produce an $F_1$ male offspring with dynamics given by $ \left[\begin{smallmatrix} 0.5(a_1 + \bar a_1) & 0.5(a_2 + \bar a_2) \\ \bar X_1 & \bar X_2 \end{smallmatrix}\right]$. If both genes resided on the autosomes this system would only be possible in an $F_2$ cross.}

As far as we know, this mechanism for Haldane's rule is novel and distinct from the currently favored ``dominance'', ``faster-X'', and ``faster-male'' theories \citep{orr1997haldane, coyne1998evolutionary}. In fact, this mechanism makes different predictions than the dominance theory as formulated by \citet{turelli1995dominance}. \citet{turelli1995dominance} employ a population genetic model which strictly requires alleles causing hybrid incompatibility on the X chromosome not to exhibit additive effects, that is the \emph{dominance coefficient} (typically $h$ in classic population genetics) must not be equal to $\frac{1}{2}$. In their model, if alleles are strictly additive, that is neither dominant nor recessive, then male and female $F_1$ hybrids will drop in fitness at the exact same rates, and therefore defy Haldane's rule. Furthermore, the dominance theory struggles with dosage compensation and X-inactivation \citep{watson2012haldane}; the present model, however, complies with Haldane's rule with X-inactivation explicitly included.

>>>>

\begin{abstract}
It is known that even if a species' phenotype has remained unchanged over evolutionary time,
the underlying mechanism may have changed,
since distinct molecular pathways can realize identical phenotypes.
In this paper, we use quantitative genetics and linear systems theory to study
how a gene network underlying a conserved phenotype
evolves as genetic drift of small mutational tweaks to the molecular pathways
cause a population to explore the set of mechanisms with identical phenotypes.
In this setting we treat organisms as ``black boxes'' for which the environment provides input
and the phenotype is the output,
and there exists an exact characterization of the set of all mechanisms that give the same input--output relationship. 
We show that in this situation, there is never a unique architecture for any phenotype
and that the evolutionary exploration of these distinct and mutationally connected mechanisms
can lead to the rapid accumulation of hybrid incompatibilities between allopatric populations.
We estimate that in reasonably numerous species,
this process could thus the formation of new species
over a number of generations proportional to the effective population size.
This model also provides a natural explanation for Haldane's rule
that appears mechanistically distinct from other proposed mechanisms.
\end{abstract}

@@

\paragraph{Haldane's rule}
This model naturally predicts
Haldane's rule, the observation that if only one hybrid sex is sterile or inviable 
it is likely the heterogametic sex (e.g., the male in XY sex determination systems) \citep{haldane1922sex}.
For example, consider an XY species with a two-gene network where the first gene resides on an autosome and the second gene on the X chromosome.
A male whose pair of haplotypes are 
$\left( \left[ \begin{smallmatrix} A_1 & A_2 \\ \cdot & \cdot \end{smallmatrix} \right], \left[ \begin{smallmatrix} A_1 & A_2 \\ X_1 & X_2 \end{smallmatrix} \right] \right)$
has phenotype determined by $A = \left[ \begin{smallmatrix} A_1 & A_2 \\ X_1/2 & X_2/2 \end{smallmatrix} \right]$,
while a female homozygous for the haplotype
$\left[\begin{smallmatrix} \bar A_1 & \bar A_2 \\ \bar X_1 & \bar X_2 \end{smallmatrix}\right]$,
thanks to X-inactivation, has phenotype determined by
$A = \left[\begin{smallmatrix} \bar A_1 & \bar A_2 \\ \bar X_1/2 & \bar X_2/2 \end{smallmatrix}\right]$.
An $F_1$ male offspring of these two will have phenotype determined by
$\frac{1}{2} \left[\begin{smallmatrix} A_1 + \bar A_1 & A_2 + \bar A_2 \\ \bar X_1 & \bar X_2 \end{smallmatrix}\right]$.
If both genes resided on the autosomes this system would only be possible in an $F_2$ cross.
More generally, 
if the regulatory coefficients for a system are shared between the sex and one or more autosomal chromosomes, 
$F_1$ males are effectively equivalent to purely autosomal-system $F_2$s.

This mechanism is apparently distinct from the currently favored ``dominance'', ``faster-X'', and ``faster-male'' theories \citep{orr1997haldane, coyne1998evolutionary} to explain Haldane's rule,
and makes several different predictions than the dominance theory as formulated by \citet{turelli1995dominance}.
\plr{what are they?}
\citet{turelli1995dominance} employ a population genetic model which strictly requires alleles causing hybrid incompatibility on the X chromosome not to exhibit additive effects, that is the \emph{dominance coefficient} (typically $h$ in classic population genetics) must not be equal to $\frac{1}{2}$. In their model, if alleles are strictly additive, that is neither dominant nor recessive, then male and female $F_1$ hybrids will drop in fitness at the exact same rates, and therefore defy Haldane's rule. 
Furthermore, the dominance theory struggles with dosage compensation and X-inactivation \citep{watson2012haldane}; the present model, however, complies with Haldane's rule with X-inactivation explicitly included.
